\LARGE \bf Portfolio Optimization Based on a Complex Networks Model\\ \normalsize TODO: propose a better title

### 2017-05-07 tex/article.tex
Asset extraction

  A new portfolio optimization algorithm is presented that is based on statistical arbitrage and potential method for determining the preference flow.
  At each time step a graph that represents preference relations among financial assets (i.e., if connection exists from asset A to asset B then A is preferred over B) is constructed, using the modified version of statistical arbitrage.
  Then, the preference flow is calculated, using the potential method\cite{caklovic}, from which preferred assets are selected into the portfolio for that time step.

  Method has been tested on dataset XY and... (TODO: what happened)

@@

  The task of portfolio optimization is to try to enhance various criteria, which most of the time include maximization of expected return and minimization of deviation...

  The approach in this paper relies on abrupt deviations in relations between prices in some observed set of assets, so in general it performs better where there is larger number of assets.

@@

@@

  Parameters of the algorithm are: $T$ - number of days, and $p$ - deviation threshold.

@@

  \begin{equation}
    \label{eq:thresh}
    \left| c_{i,j}^{(t)} - m_{i,j}^{(t-1)} \right| > p \cdot d_{i,j}^{(t - 1)},
  \end{equation}
  i.e. current log price difference is at least $p$ deviations distant from mean value of the past time window.

@@

  \begin{equation}
    \label{eq:weight}
    w_{i,j}^{(t)} = \left(c_{i,j}^{(t)} - \left. m_{i,j}^{(t-1)}\right) \middle/ d_{i,j}^{(t-1)} \right..
  \end{equation}

  Thus, it is possible to create a relatively sparse graph for each time step $t \in T..D-1$.
  At some time steps it is possible that the graph could be empty, if it is the case that no pair $(i,j)$ satisfies (\ref{eq:thresh}).
  Lower values of parameter $p$ yield denser graphs.

@@

  From previously obtained graph it is possible to tell which pair has the highest preference.
  However it is still not possible to tell which asset is the most preferable one, or obtain the measure of preference for individual assets, directly.
  For this we need to calculate some sort of preference flow, as if it were water going through pipes of widths corresponding to edge weights, from one asset to another.
  A quick way to do this is by using the potential method as described in \cite{caklovic}.

>>>>

  A new algorithm for portfolio optimization is presented which is based on statistical arbitrage, with potential method used to obtain the most preferred assets.
  A graph that represents preference relations among financial assets (i.e., if an edge exists going from asset A to asset B, then A is preferred over B) is constructed at each time step a, using the modified version of statistical arbitrage.
  Then, the preference flow of each asset is calculated, using the potential method\cite{caklovic}, from which the most preferred assets are selected into the portfolio for each time step.

  Method has been tested on dataset XY, and\dots (TODO: what happened)

@@

  The task of portfolio optimization is to try to enhance various criteria, which most of the time include maximization of expected return and minimization of deviation\dots

  The approach that is taken in this paper relies on finding abrupt deviations of price relations between in the observed set of assets.
  Algorithm works on pairs of assets, looking for such uncommon deviations, so generally it is expected to perform better where there is larger number of assets.
%  Each asset is compared to all other assets while looking for such specific deviations, so, in general, this algorithm performs better where there is larger number of assets.

@@

  Include measure of inconsistency?

@@

  Parameters of the algorithm are:
  \begin{itemize}
    \item $T$ - length of the past time window
    \item $\alpha$ - the deviation threshold
  \end{itemize}

@@

  \begin{equation}
    \label{eq:thresh}
    \left| c_{i,j}^{(t)} - m_{i,j}^{(t-1)} \right| > \alpha \cdot d_{i,j}^{(t - 1)},
  \end{equation}
  i.e. current log price difference is at least $\alpha$ deviations distant from mean value of the past time window.

@@

  \begin{equation}
    \label{eq:weight}
    w_{i,j}^{(t)} = \left. \left(c_{i,j}^{(t)} - m_{i,j}^{(t-1)}\right) \middle/ d_{i,j}^{(t-1)} \right..
  \end{equation}

  Thus, it is possible to create a relatively sparse graph for each time step $t \in T..D-1$.
  At some time steps it is possible that the graph could be empty, if it is the case that no pair $(i,j)$ satisfies (\ref{eq:thresh}).
  Lower values of parameter $\alpha$ yield denser graphs.

@@

  From previously obtained graph it is possible to tell which pair of assets has the highest preference flow.
  However it is not yet possible to directly tell which is the most preferrable asset, or obtain the measure of preference for individual assets.
  For this we need to calculate some sort of potential for each node in graph, which corresponds to difference in amount of flow going in and out of the node.
  A quick way to do this is by using the potential method as described in \cite{caklovic}.
  The summary of the method is as follows:
  \begin{enumerate}
    \item Let $\mathbf{A}$ be the incidence matrix of previously obtained graph. $\mathbf{A}$ has following properties:
    \begin{enumerate}
      \item each row corresponds to an edge in the graph, and each column to a node,
      \item for every edge in the graph going from node $i$ to node $j$, there is a corresponding row that has $-1$ and $1$ in columns that correspond to nodes $i$ and $j$, respectively,
      \item the remainder of elements in the matrix are zeros.
    \end{enumerate}
    \item Let $\mathbf{F}$ be column vector that contains edge weights, and order of the edges is consistent with order of the edges in $\mathbf{A}$.

    \item Let $\mathbf{X}$ be column vector that contains potentials of each node, and order of the nodes is consistent with order of the nodes in $\mathbf{A}$.

    \item Now $\mathbf{A}$, $\mathbf{X}$, and $\mathbf{F}$ should satisfy the equation
    \begin{equation}
      \label{eq:flowover}
      \mathbf{A} \mathbf{X} = \mathbf{F},
    \end{equation}
    meaning that the difference between potential of any two nodes should result result in weight of the edge between them.
    Most of the time this represents an overdetermined system since $\mathbf{A}$ has more rows than columns, so we convert it to a least squares problem:
    \begin{equation}
      \label{eq:flow}
      \begin{gathered}
      \min_{\mathbf{X}} \left\{ \left \lVert \mathbf{A} \mathbf{X} - \mathbf{F}\textsl{} \right \rVert ^ 2 \right\} \Rightarrow
      \frac{\partial}{\partial \mathbf{X}} \left \lVert \mathbf{A} \mathbf{X} - \mathbf{F}\textsl{} \right \rVert ^ 2 = \mathbf{0},
      \end{gathered}
    \end{equation}
    from which we obtain the equation:
    \begin{equation}
      \mathbf{A}^{T} \mathbf{A} \mathbf{X} = \mathbf{A}^{T} \mathbf{F}.
    \end{equation}
    Following constraint is also added:
    \begin{equation}
      \label{eq:sumiszero}
      \underbrace{\begin{bmatrix} 1 & 1 & \cdots & 1 \end{bmatrix}}_{N} \cdot \mathbf{X} = 0
    \end{equation}
    to ensure that total amounts of positive and negative potential are equal.

    \item Joining the previous two equations together by adding (\ref{eq:sumiszero}) to each row of (\ref{eq:flow}) results in:
    \begin{equation}
      \begin{aligned}
        \mathbf{A}^{T} \mathbf{A} \mathbf{X} + \mathbf{J} \mathbf{X} &= \mathbf{A}^{T} \mathbf{F}\\
        \left[\mathbf{A}^{T} \mathbf{A} + \mathbf{J} \right] \mathbf{X} &= \mathbf{A}^{T} \mathbf{F},
      \end{aligned}
    \end{equation}
    where $\mathbf{J}$ is a matrix of ones with same dimension as $\mathbf{A}^{T} \mathbf{A}$.
    Finally, solving for $\mathbf{X}$ gives us:
    \begin{equation}
      \mathbf{X} = \left[\mathbf{A}^T \mathbf{A} + \mathbf{J} \right]^{-1} \mathbf{A}^T \mathbf{F}.
    \end{equation}
  \end{enumerate}
